On the propriety of restricted Boltzmann machines

### 2016-12-01 writing/draft.Rmd
nordman and vardeman final edits

For a RBM model with enough hidden variables, it has been shown that any distribution for the visibles can be approximated arbitrarily well [@le2008representational; @montufar2011refinements; and @montufar2011expressive]. However, the empirical distribution of a training set of vectors of visibles is the best fitting model for observed cell data. As a consequence, we find that any fully principled fitting method based on the likelihood for RBM with enough hidden variables will simply seek to reproduce the (discrete) empirical distribution of a training set. Thus, there can be no "smoothed distribution" achieved in fitting a RBM model of sufficient size with a rigorous likelihood-based method. We are therefore led to be skeptical that any model built using these structures (like a deep Boltzmann machine) can achieve prediction or inference in a principled way without intentionally limiting the flexibility of the fitted model.

@@

where $\boldsymbol \theta \equiv (\theta_{11}, \dots, \theta_{1\nh}, \dots, \theta_{\nv 1}, \dots, \theta_{\nv \nh}, \theta_{v_1}, \dots, \theta_{v_\nv}, \theta_{h_1}, \dots, \theta_{h_\nh}) \in \mathbb{R}^{\nv + \nh + \nv*\nh}$ denotes the vector of model parameters and the denominator 
$$\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{C}^{\nh+\nv}}\exp\left(\sum_{i = 1}^\nv \sum_{j=1}^\nh \theta_{ij} v_i h_j + \sum_{i = 1}^\nv\theta_{v_i} v_i + \sum_{j = 1}^\nh\theta_{h_j} h_j\right)
$$ 
is the normalizing function that ensures the probabilities (\ref{eqn:pmf}) sum to one. For $\boldsymbol x = (h_1, \dots, h_\nh, v_1, \dots, v_\nh) \in \mathcal{C}^{\nv + \nh}$ and 
\begin{align}
\boldsymbol t(\boldsymbol x) &= (h_1, \dots, h_\nh, v_1, \dots, v_\nv, v_1 h_1, \dots, v_\nv h_\nh) \in \mathcal{C}^{\nh + \nv + \nh*\nv}, \label{eqn:t}
\end{align}
let $\mathcal{T} = \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nv + \nh}\} \subset \mathbb{R}^{\nv + \nh + \nv * \nh}$ be the set of possible values for the vector of variables needed to compute probabilities (\ref{eqn:pmf}) in the model and write $Q_{\boldsymbol \theta}(\boldsymbol x) = \sum\limits_{i = 1}^\nh\sum\limits_{j=1}^\nv \theta_{ij} h_i v_j + \sum\limits_{i = 1}^\nh\theta_{hi} h_i + \sum\limits_{j = 1}^\nv\theta_{vj} v_j$ for the "neg-potential" function. The RBM model is parameterized by $\boldsymbol \theta$ containing two types of parameters, main effects and interaction effects. The main effects parameters ($\{\theta_{v_i}, \theta_{h_j}\}_{\substack{i = 1, \dots, \nv,\\j = 1, \dots, \nh}}$) weight the values of the visible $v_i$ and hidden $h_j$ nodes in probabilities (\ref{eqn:pmf}) and the interaction effect parameters ($\theta_{ij}$) weight the values of the connections $v_i h_j$, or dependencies, between hidden and visible layers.

Due to the potential size of the model, the normalizing constant $\gamma(\boldsymbol \theta)$ can be practically impossible to calculate, making simple estimation of the model parameter vector problematic. A kind of Gibbs sampling can be tried (due to conditional architecture of the RBM, i.e. visibles given hiddens or vice verse). Specifically, the conditional independence of nodes in each layer (given those nodes in the other layer) allows for stepwise simulation of both hidden layers and model parameters (e.g., contrastive divergence of @hinton2002training or Bayes methods in Section \ref{model-fitting}).

@@

\begin{defn}[S-unstable RBM]
Let $\boldsymbol \theta_\nv \in \mathbb{R}^{\nv + \nh +  \nh*\nv}, \nv \ge 1$, be an element of a sequence of RBM parameters where the number of hiddens, $\nh \equiv \nh(\nv) \ge 1$, can be an arbitrary function of the number $\nv$ of visibles. A RBM model formulation is \emph{Schweinberger-unstable} or \emph{S-unstable} if
\begin{align*}
\lim\limits_{\nv \rightarrow \infty} \frac{1}{\nv} \text{ELPR}(\boldsymbol \theta_\nv) = \infty.
\end{align*}
\end{defn}
In other words, given any $C > 0$, there exists an integer $n_C > 0$ so that $\frac{1}{\nv}\text{ELPR}(\boldsymbol \theta_\nv) > C$ for all $\nv \ge n_C$.

@@

\begin{prop}
\label{prop:instab}
Let $C > 0$ and let $\text{ELPR}(\boldsymbol \theta_\nv)$ be as in (\ref{eqn:elpr}) for an integer $\nv \ge 1$. If $\frac{1}{\nv}\text{ELPR}(\boldsymbol \theta_\nv) > C$, then $\Delta(\boldsymbol \theta_\nv) > C$.
\end{prop}

@@

In other words, S-unstable RBM model sequences are guaranteed to stack up all probability on an arbitrarily narrow set of outcomes for visibles. Proofs of Propositions \ref{prop:instab}-\ref{prop:degen} appear in @kaplan2016note. These findings also have counterparts in results in @schweinberger2011instability, but unlike results there, we do not limit consideration to exponential family forms with a fixed number of parameters.

### Uninterpretability
For spatial Markov models, @kaiser2007statistical defines a measure of model impropriety he calls *uninterpretability*, which is characterized by dependence parameters in a model being so extreme that marginal mean-structures fail to hold as anticipated by a model statement. We adapt this notion to RBM models as follows. Note that in a RBM, the parameters $\theta_{v_1}, \dots, \theta_{v_\nv}$ and $\theta_{h_1}, \dots, \theta_{h_\nh}$ are naturally associated with main effects of visible and hidden variables and can be interpreted as (logit functions of) means for variables $V_1, \dots, V_\nv, H_1, \dots, H_\nh$ in a model with no interaction parameters, $\theta_{ij} = 0, i = 1, \dots, \nv, j = 1, \dots, \nh$. That is, with no interaction parameters, we have from (\ref{eqn:pmf}) that
$$
P_\theta(V_i=1) \propto e^{\theta_{v_i}} \quad \text{ and } \quad P_\theta(H_j=1) \propto e^{\theta_{h_j}}, \quad  i=1,\ldots,\nv,j=1,\ldots,\nh
$$
so that, for example, $\text{logit}(P_\theta(V_i=1))=\theta_{v_i}$ (or $2 \theta_{v_i}$) under the coding $\mathcal{C}=\{0,1\}$ (or $\{-1,1\}$). Hence, these main effect parameters have a clear interpretation under an independence model (one with $\theta_{ij} = 0$) but this interpretation can break down as interaction parameters increase in magnitude relative to the size of the main effects. In such cases, the main effect parameters $\theta_{v_1}$ and $\theta_{h_j}$ are no longer interpretable as originally intended in the models (statements of marginal means) and the dependence parameters are so large as to dominate the entire model probability structure (also destroying the model interpretation of dependence as local conditional modifications of an overall marginal mean structure). To assess which parameter values $\boldsymbol \theta$ may cause difficulties in interpretation, we use the difference $\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right]$ between two model expectations: E$\left[\boldsymbol X | \boldsymbol \theta\right]$ at $\boldsymbol \theta$ and expectations E$\left[\boldsymbol X | \boldsymbol \theta^* \right]$ where $\boldsymbol \theta^*$ matches $\boldsymbol \theta$ for all main effects but otherwise has $\theta_{ij} = 0$ for $i = 1, \dots, \nv, j = 1, \dots, \nh$. Hence, $\boldsymbol \theta^*$ and $\boldsymbol \theta$ have the same main effects but $\boldsymbol \theta^*$ has $0$ dependence parameters. Uninterpretability is then avoided at a parametric specification $\boldsymbol \theta$ if the model expected value at $\boldsymbol \theta$ is not very different from the corresponding model expectation under independence. Using this, it is possible to investigate what parametric conditions lead to uninterpretability in a model versus those that guarantee interpretable models. If $||\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \boldsymbol \theta^*\right]||$ is large, then the RBM model with parameter vector $\boldsymbol \theta$ is said to be uninterpetable. The quantities to compare in the RBM case are

@@

For this small model, we are able to investigate the symptoms of model impropriety, beginning with near-degeneracy. To this end, recall from Section \ref{near-degeneracy} that one characterization requires consideration of the convex hull of possible values of statistics $\boldsymbol t(\boldsymbol x)$,
$$\mathcal{T} = \{\boldsymbol t(\boldsymbol x): \boldsymbol x = (v_1, h_1) \in \{0,1\}^2\} \equiv \{(v_1, h_1, v_1 h_1): v_1, h_1 \in \{0,1\}\}$$ 
appearing in the RBM probabilities for this model. As this set is in three dimensions, we are able to explicitly illustrate the shape of boundary of the convex hull of $\mathcal{T}$ and explore the behavior of the mean vector $\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol x)$ as a function of the parameter vector $\boldsymbol \theta$. Figure \ref{fig:toyhull} shows the convex hull of our "statistic space," $\mathcal{T} \subset \{0,1\}^3$, for this toy problem from two perspectives (enclosed by the unit cube $[0,1]^3$, representing the convex hull of $\{0,1\}^3$ space). In this small model, note that the convex hull of $\mathcal{T}$ does not fill the unrestricted hull of $\{0,1\}^3$ because of the relationship between the elements of $\mathcal{T} = \{(v_1, h_1, v_1 h_1 : v_1, h_1 \in \{0, 1\} \}$ (i.e. $v_1 h_1 = 1$ only if $v_1 = h_1 = 1$).

@@

We can compute the mean vector for $\boldsymbol t(\boldsymbol x)$ as a function of the model parameters as 
\begin{align}
\boldsymbol \mu(\boldsymbol \theta) &= \text{E}_{\boldsymbol \theta}\left[ \boldsymbol t(\boldsymbol X) \right] =  \sum\limits_{\boldsymbol x = (v_1, h_1) \in \mathcal{X} = \{0,1\}^2} \left\{ t(x) \frac{\exp\left( \theta_{11} h_1 v_1 + \theta_{h1} h_1 + \theta_{v1} v_1\right)}{\gamma(\boldsymbol \theta)} \right\}
= \left[
\begin{matrix}
\frac{\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)}{\gamma(\boldsymbol \theta)} \\
\frac{\exp\left(\theta_{h_1}\right) + \exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)}{\gamma(\boldsymbol \theta)} \\
\frac{\exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)}{\gamma(\boldsymbol \theta)} \\
\end{matrix} \right]
\end{align}
where $\gamma(\boldsymbol \theta) = \sum\limits_{h_1 = 0}^1\sum\limits_{v_1 = 0}^1 \exp(\theta_{11} h_1 v_1 + \theta_{h_1}h_1 + \theta_{v_1}v_1)$. The three parametric coordinate functions of $\boldsymbol \mu(\boldsymbol \theta)$ can be represented as in Figure \ref{fig:degen_toy}. (Contour plots for three coordinate functions are shown in columns for various values of $\theta_{11}$.) In examining these, we see that as values of coordinates of $\boldsymbol \theta$ grow larger in magnitude, at least one mean function for the entries of $\boldsymbol t(\boldsymbol x)$ is approaching a value 0 or 1, forcing $\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol x)$ to be near to the boundary of the convex hull of $\mathcal{T}$, as a sign of model near-degeneracy. Thus, for a very small example we can see the relationship between values of $\boldsymbol \theta$ and model degeneracy.

@@

```{r uninterp, fig.cap="\\label{fig:uninterp}The absolute difference between coordinates of model expectations, E$\\left[\\boldsymbol X | \\boldsymbol \\theta\\right]$, and expectations given independence, E$\\left[\\boldsymbol X | \\boldsymbol \\theta^* \\right ]$ for a RBM with one visible and one hidden node. As an indicator of uninterpretability, note that differences in expectations increase as the dependence parameter $\theta_{11}$ deviates from zero.", fig.height = 4, fig.align='center', message=FALSE, cache=TRUE, warning=FALSE}
expand.grid(theta_h1 = seq(-5, 5, length.out = 21),
            theta_v1 = seq(-5, 5, length.out = 21),
            theta_11 = seq(-5, 5, length.out = 5)) %>%
  data.frame() %>%
  group_by(theta_v1, theta_h1, theta_11) %>%
  do(data.frame(gamma = sum(exp(possibles %*% t(data.matrix(.)))),
                mu_v1_dep = possibles[,"v1"] %*% exp(possibles %*% t(data.matrix(.))),
                mu_h1_dep = possibles[,"h1"] %*% exp(possibles %*% t(data.matrix(.))))) %>%
  mutate_each(funs(./gamma), starts_with("mu")) %>%
  select(-gamma) -> exp_dep

@@

In addition to the argument that the $-1/1$ data encoding mitigates some perceived prevalence of near-degeneracy it also has the benefit of providing a guaranteed-to-be non-degenerate model at $\boldsymbol \theta = \boldsymbol 0 \in \mathbb{R}^{\nh + \nv + \nh*\nv}$, where the zero vector then serves as the natural center of the parameter space and induces the simplest possible model properties for the RBM (i.e., at $\boldsymbol \theta = \boldsymbol 0$, all variables are independent and visible variables are independent and uniformly distributed on $\{-1,1\}^\nv$). The proof of this and further exploration of the equivalence of the $\boldsymbol \theta$ parameterization of the RBM model class and parameterization by $\boldsymbol \mu(\boldsymbol \theta)$ is in the on-line supplementary materials. Hence while from some computing perspectives $0/1$ coding might seem most natural, the $-1/1$ coding is far more convenient and interpretable from the point of view of statistical modeling, where it makes parameters simply interpreted in terms of symmetrically defined main effects and interactions. In light of all of these matters we will henceforth employ the $-1/1$ coding.

@@

However, computational concerns are not the only issues with fitting a RBM using ML. In addition, a RBM model, with the appropriate choice of parameters and number of hiddens, has the potential to re-create any distribution for the data (i.e., reproduce any specification of cell probabilities for the binary data outcomes). For example, @montufar2011refinements show that any distribution on $\{0, 1\}^{\nv}$ can be approximated arbitrarily well by a RBM with $2^{\nv}/2 - 1$ hidden units. To illustrate, we provide a small example.

@@

cbind(tbl[1:8,], tbl[9:16,], tbl[17:24,]) %>%
  kable(caption = "\\label{tab:theta}Parameters used to fit a test case with $V = H = 4$. This parameter vector was chosen as a sampled value of $\\boldsymbol \\theta$ that was not near the convex hull of the sufficient statistics for a grid point in Figure \\ref{fig:degen_plots} with $< 5$\\% near-degeneracy.", align = rep(c("l", "r"), times = 3))
```

We considered a test case with $\nv = \nh = 4$ and parameters given in Table \ref{tab:theta}. This parameter vector was chosen as a sampled value of $\boldsymbol \theta$ that was not near the convex hull of the sufficient statistics for a grid point in Figure \ref{fig:degen_plots} with $< 5$\% near-degeneracy. We simulated $n = 5,000$ realizations of visibles as a training set and fit the RBM using three Bayes methodologies. These involved the following:

@@

The effective sample sizes for a chain of length $M = 1000$ for inference about each of the $2^4 = 16$ model probabilities are presented in Table \ref{tab:m_eff}. These range from  $`r round(min(M_eff_tmp[, "BwTNML"]), 2)`$ to $`r round(max(M_eff_tmp[, "BwTNML"]), 2)`$ for BwTNML, while BwTNLV only yields between $`r round(min(M_eff_tmp[, "BwTNLV"]), 2)`$ and $`r round(max(M_eff_tmp[, "BwTNLV"]), 2)`$ effective draws. Thus, BwTNLV would require at least $`r round(min(M_eff_tmp[, "BwTNML"])/min(M_eff_tmp[, "BwTNLV"]), 1)`$ times more iterations to be run of the MCMC chain in order to achieve the same amount of effective information about the posterior distribution. For this reason, consistent with the ACF results in Figure \ref{fig:acf}, BwTNLV does not seem to be an effective method for fitting the RBM if computing resources are at all limited.

>>>>

For a RBM model with enough hidden variables, it has been shown that any distribution for the visibles can be approximated arbitrarily well [@le2008representational; @montufar2011refinements; and @montufar2011expressive]. However, the empirical distribution of a training set of vectors of visibles is the best fitting model for observed cell data. As a consequence, we find that any fully principled fitting method based on the likelihood for a RBM with enough hidden variables will simply seek to reproduce the (discrete) empirical distribution of a training set. Thus, there can be no "smoothed distribution" achieved in fitting a RBM model of sufficient size with a rigorous likelihood-based method. We are therefore led to be skeptical that any model built using these structures (like a deep Boltzmann machine) can achieve useful prediction or inference in a principled way without intentionally limiting the flexibility of the fitted model.

@@

where $\boldsymbol \theta \equiv (\theta_{11}, \dots, \theta_{1\nh}, \dots, \theta_{\nv 1}, \dots, \theta_{\nv \nh}, \theta_{v_1}, \dots, \theta_{v_\nv}, \theta_{h_1}, \dots, \theta_{h_\nh}) \in \mathbb{R}^{\nv + \nh + \nv*\nh}$ denotes the vector of model parameters and the denominator 
$$\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{C}^{\nh+\nv}}\exp\left(\sum_{i = 1}^\nv \sum_{j=1}^\nh \theta_{ij} v_i h_j + \sum_{i = 1}^\nv\theta_{v_i} v_i + \sum_{j = 1}^\nh\theta_{h_j} h_j\right)
$$ 
is the normalizing function that ensures the probabilities (\ref{eqn:pmf}) sum to one. For $\boldsymbol x = (h_1, \dots, h_\nh, v_1, \dots, v_\nh) \in \mathcal{C}^{\nv + \nh}$ and 
\begin{align}
\boldsymbol t(\boldsymbol x) &= (h_1, \dots, h_\nh, v_1, \dots, v_\nv, v_1 h_1, \dots, v_\nv h_\nh) \in \mathcal{C}^{\nh + \nv + \nh*\nv}, \label{eqn:t}
\end{align}
let $\mathcal{T} = \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nv + \nh}\} \subset \mathbb{R}^{\nv + \nh + \nv * \nh}$ be the set of possible values for the vector of variables needed to compute probabilities (\ref{eqn:pmf}) in the model, and write $Q_{\boldsymbol \theta}(\boldsymbol x) = \sum\limits_{i = 1}^\nh\sum\limits_{j=1}^\nv \theta_{ij} h_i v_j + \sum\limits_{i = 1}^\nh\theta_{hi} h_i + \sum\limits_{j = 1}^\nv\theta_{vj} v_j$ for the "neg-potential" function. The RBM model is parameterized by $\boldsymbol \theta$ containing two types of parameters, main effects and interaction effects. The main effects parameters ($\{\theta_{v_i}, \theta_{h_j}\}_{\substack{i = 1, \dots, \nv,\\j = 1, \dots, \nh}}$) weight the values of the visible $v_i$ and hidden $h_j$ nodes in probabilities (\ref{eqn:pmf}) and the interaction effect parameters ($\theta_{ij}$) weight the values of the connections $v_i h_j$, or dependencies, between hidden and visible layers.

Due to the potential size of the model, the normalizing constant $\gamma(\boldsymbol \theta)$ can be practically impossible to calculate, making simple estimation of the model parameter vector problematic. A kind of Gibbs sampling can be tried (due to the conditional architecture of the RBM, i.e. visibles given hiddens or vice verse). Specifically, the conditional independence of nodes in each layer (given those nodes in the other layer) allows for stepwise simulation of both hidden layers and model parameters (e.g., see the contrastive divergence of @hinton2002training or Bayes methods in Section \ref{model-fitting}).

@@

\begin{defn}[S-unstable RBM]
Let $\boldsymbol \theta_\nv \in \mathbb{R}^{\nv + \nh +  \nh*\nv}, \nv \ge 1$, be an element of a sequence of RBM parameters where the number of hiddens, $\nh \equiv \nh(\nv) \ge 1$, can be an arbitrary function of the number $\nv$ of visibles. A RBM model formulation is \emph{Schweinberger-unstable} or \emph{S-unstable} if
\begin{align*}
\lim\limits_{\nv \rightarrow \infty} \frac{1}{\nv} \text{ELPR}(\boldsymbol \theta_\nv) = \infty.
\end{align*}
\end{defn}
In other words, given any $c > 0$, there exists an integer $n_c > 0$ so that $\frac{1}{\nv}\text{ELPR}(\boldsymbol \theta_\nv) > c$ for all $\nv \ge n_c$.

@@

\begin{prop}
\label{prop:instab}
Let $c > 0$ and let $\text{ELPR}(\boldsymbol \theta_\nv)$ be as in (\ref{eqn:elpr}) for an integer $\nv \ge 1$. If $\frac{1}{\nv}\text{ELPR}(\boldsymbol \theta_\nv) > c$, then $\Delta(\boldsymbol \theta_\nv) > c$.
\end{prop}

@@

In other words, S-unstable RBM model sequences are guaranteed to stack up all probability on an arbitrarily narrow set of outcomes for visibles. Proofs of Propositions \ref{prop:instab} and \ref{prop:degen} appear in @kaplan2016note. These findings also have counterparts in results in @schweinberger2011instability, but unlike results there, we do not limit consideration to exponential family forms with a fixed number of parameters.

### Uninterpretability
For spatial Markov models, @kaiser2007statistical defines a measure of model impropriety he calls *uninterpretability*, which is characterized by dependence parameters in a model being so extreme that marginal mean-structures fail to hold as anticipated by a model statement. We adapt this notion to RBM models as follows. Note that in a RBM, the parameters $\theta_{v_1}, \dots, \theta_{v_\nv}$ and $\theta_{h_1}, \dots, \theta_{h_\nh}$ are naturally associated with main effects of visible and hidden variables and can be interpreted as (logit functions of) means for variables $V_1, \dots, V_\nv, H_1, \dots, H_\nh$ in a model with no interaction parameters, $\theta_{ij} = 0, i = 1, \dots, \nv, j = 1, \dots, \nh$. That is, with no interaction parameters, we have from (\ref{eqn:pmf}) that
$$
P_\theta(V_i=1) \propto e^{\theta_{v_i}} \quad \text{ and } \quad P_\theta(H_j=1) \propto e^{\theta_{h_j}}, \quad  i=1,\ldots,\nv,j=1,\ldots,\nh
$$
so that, for example, $\text{logit}(P_\theta(V_i=1))=\theta_{v_i}$ (or $2 \theta_{v_i}$) under the coding $\mathcal{C}=\{0,1\}$ (or $\{-1,1\}$). Hence, these main effect parameters have a clear interpretation under an independence model (one with $\theta_{ij} = 0$) but this interpretation can break down as interaction parameters increase in magnitude relative to the size of the main effects. In such cases, the main effect parameters $\theta_{v_1}$ and $\theta_{h_j}$ are no longer interpretable as originally intended in the models (statements of marginal means) and the dependence parameters are so large as to dominate the entire model probability structure (also destroying the model interpretation of dependence as local conditional modifications of an overall marginal mean structure). To assess which parameter values $\boldsymbol \theta$ may cause difficulties in interpretation, we use the difference $\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right]$ between two model expectations: E$\left[\boldsymbol X | \boldsymbol \theta\right]$ at $\boldsymbol \theta$ and expectations E$\left[\boldsymbol X | \boldsymbol \theta^* \right]$ where $\boldsymbol \theta^*$ matches $\boldsymbol \theta$ for all main effects but otherwise has $\theta_{ij} = 0$ for $i = 1, \dots, \nv, j = 1, \dots, \nh$. Hence, $\boldsymbol \theta^*$ and $\boldsymbol \theta$ have the same main effects but $\boldsymbol \theta^*$ has $0$ dependence parameters. Uninterpretability is then avoided at a parametric specification $\boldsymbol \theta$ if the model expected value at $\boldsymbol \theta$ is not very different from the corresponding model expectation under independence. Using this, it is possible to investigate what parametric conditions lead to uninterpretability in a model versus those that guarantee interpretable models. If $\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] - \text{E}\left[\boldsymbol X | \boldsymbol \theta^*\right]$ is large, then the RBM model with parameter vector $\boldsymbol \theta$ is said to be uninterpetable. The quantities to compare in the RBM case are

@@

For this small model, we are able to investigate the symptoms of model impropriety, beginning with near-degeneracy. To this end, recall from Section \ref{near-degeneracy} that one characterization requires consideration of the convex hull of possible values of statistics $\boldsymbol t(\boldsymbol x)$,
$$\mathcal{T} = \{\boldsymbol t(\boldsymbol x): \boldsymbol x = (v_1, h_1) \in \{0,1\}^2\} \equiv \{(v_1, h_1, v_1 h_1): v_1, h_1 \in \{0,1\}\}$$ 
appearing in the RBM probabilities for this model. As this set is in three dimensions, we are able to explicitly illustrate the shape of boundary of the convex hull of $\mathcal{T}$ and explore the behavior of the mean vector $\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol x)$ as a function of the parameter vector $\boldsymbol \theta$. Figure \ref{fig:toyhull} shows the convex hull of our "statistic space," $\mathcal{T} \subset \{0,1\}^3$, for this toy problem from two perspectives (enclosed by the unit cube $[0,1]^3$, the convex hull of $\{0,1\}^3$). In this small model, note that the convex hull of $\mathcal{T}$ does not fill the unrestricted hull of $\{0,1\}^3$ because of the relationship between the elements of $\mathcal{T} = \{(v_1, h_1, v_1 h_1 : v_1, h_1 \in \{0, 1\} \}$ (i.e. $v_1 h_1 = 1$ only if $v_1 = h_1 = 1$).

@@

We can compute the mean vector for $\boldsymbol t(\boldsymbol x)$ as a function of the model parameters as 
\begin{align}
\boldsymbol \mu(\boldsymbol \theta) &= \text{E}_{\boldsymbol \theta}\left[ \boldsymbol t(\boldsymbol X) \right] =  \sum\limits_{\boldsymbol x = (v_1, h_1) \in \mathcal{X} = \{0,1\}^2} \left\{ t(x) \frac{\exp\left( \theta_{11} h_1 v_1 + \theta_{h1} h_1 + \theta_{v1} v_1\right)}{\gamma(\boldsymbol \theta)} \right\}
= \left[
\begin{matrix}
\frac{\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)}{\gamma(\boldsymbol \theta)} \\
\frac{\exp\left(\theta_{h_1}\right) + \exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)}{\gamma(\boldsymbol \theta)} \\
\frac{\exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)}{\gamma(\boldsymbol \theta)} \\
\end{matrix} \right]
\end{align}
where $\gamma(\boldsymbol \theta) = \sum\limits_{h_1 = 0}^1\sum\limits_{v_1 = 0}^1 \exp(\theta_{11} h_1 v_1 + \theta_{h_1}h_1 + \theta_{v_1}v_1)$. The three parametric coordinate functions of $\boldsymbol \mu(\boldsymbol \theta)$ can be represented as in Figure \ref{fig:degen_toy}. (Contour plots for three coordinate functions are shown in columns for various values of $\theta_{11}$.) In examining these, we see that as coordinates of $\boldsymbol \theta$ grow larger in magnitude, at least one mean function for the entries of $\boldsymbol t(\boldsymbol x)$ approaches a value 0 or 1, forcing $\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol x)$ to be near to the boundary of the convex hull of $\mathcal{T}$, as a sign of model near-degeneracy. Thus, for a very small example we can see the relationship between values of $\boldsymbol \theta$ and model degeneracy.

@@

```{r uninterp, fig.cap="\\label{fig:uninterp}The absolute difference between coordinates of model expectations, E$\\left[\\boldsymbol X | \\boldsymbol \\theta\\right]$, and expectations given independence, E$\\left[\\boldsymbol X | \\boldsymbol \\theta^* \\right ]$ for a RBM with one visible and one hidden node. As an indicator of uninterpretability, note that differences in expectations increase as the dependence parameter $\\theta_{11}$ deviates from zero.", fig.height = 4, fig.align='center', message=FALSE, cache=TRUE, warning=FALSE}
expand.grid(theta_h1 = seq(-5, 5, length.out = 21),
            theta_v1 = seq(-5, 5, length.out = 21),
            theta_11 = seq(-5, 5, length.out = 5)) %>%
  data.frame() %>%
  group_by(theta_v1, theta_h1, theta_11) %>%
  do(data.frame(gamma = sum(exp(possibles %*% t(data.matrix(.)))),
                mu_v1_dep = possibles[,"v1"] %*% exp(possibles %*% t(data.matrix(.))),
                mu_h1_dep = possibles[,"h1"] %*% exp(possibles %*% t(data.matrix(.))))) %>%
  mutate_each(funs(./gamma), starts_with("mu")) %>%
  select(-gamma) -> exp_dep

@@

In addition to the argument that the $-1/1$ data encoding mitigates some perceived prevalence of near-degeneracy it also has the benefit of providing a guaranteed-to-be non-degenerate model at $\boldsymbol \theta = \boldsymbol 0 \in \mathbb{R}^{\nh + \nv + \nh*\nv}$, where the zero vector then serves as the natural center of the parameter space and induces the simplest possible model properties for the RBM (i.e., at $\boldsymbol \theta = \boldsymbol 0$, all variables are independent and visible variables are independent and uniformly distributed on $\{-1,1\}^\nv$). The proof of this and further exploration of the equivalence of the $\boldsymbol \theta$ parameterization of the RBM model class and parameterization by $\boldsymbol \mu(\boldsymbol \theta)$ is in the on-line supplementary materials. Hence, while from some computing perspectives $0/1$ coding might seem most natural, the $-1/1$ coding is far more convenient and interpretable from the point of view of statistical modeling, where it makes parameters simply interpreted in terms of symmetrically defined main effects and interactions. In light of all of these matters we will henceforth employ the $-1/1$ coding.

@@

However, computational concerns are not the only issues with fitting a RBM using ML. In addition, a RBM model, with the appropriate choice of parameters and number of hiddens, has the potential to re-create any distribution for the data (i.e., reproduce any specification of cell probabilities for the binary data outcomes). For example, @montufar2011refinements show that any distribution on $\{0, 1\}^{\nv}$ can be approximated arbitrarily well by a RBM with $2^{\nv-1} - 1$ hidden units. We provide a small example that illustrates that in fact there can be many such approximations.

@@

cbind(tbl[1:8,], tbl[9:16,], tbl[17:24,]) %>%
  kable(caption = "\\label{tab:theta}Parameters used to fit a test case with $V = H = 4$. This parameter vector was chosen as a sampled value of $\\boldsymbol \\theta$ that was not near the convex hull of the sufficient statistics for a grid point in Figure \\ref{fig:degen_plots} with $< 5$\\% near-degeneracy.", align = rep(c("r", "r"), times = 3))
```

We considered a test case with $\nv = \nh = 4$ and parameters given in Table \ref{tab:theta}. This parameter vector was chosen as a sampled value of $\boldsymbol \theta$ that was not near the convex hull of the space of values of sufficient statistics for a grid point in Figure \ref{fig:degen_plots} with $< 5$\% near-degeneracy. We simulated $n = 5,000$ realizations of visibles as a training set and fit the RBM using three Bayes methodologies. These involved the following:

@@

The effective sample sizes for a chain of length $M = 1000$ for inference about each of the $2^4 = 16$ model probabilities are presented in Table \ref{tab:m_eff}. These range from  $`r round(min(M_eff_tmp[, "BwTNML"]), 2)`$ to $`r round(max(M_eff_tmp[, "BwTNML"]), 2)`$ for BwTNML, while BwTNLV only yields between $`r round(min(M_eff_tmp[, "BwTNLV"]), 2)`$ and $`r round(max(M_eff_tmp[, "BwTNLV"]), 2)`$ effective draws. Thus, BwTNLV would require at least $`r round(min(M_eff_tmp[, "BwTNML"])/min(M_eff_tmp[, "BwTNLV"]), 1)`$ times as many iterations to be run of the MCMC chain in order to achieve the same amount of effective information about the posterior distribution. For this reason, consistent with the ACF results in Figure \ref{fig:acf}, BwTNLV does not seem to be an effective method for fitting the RBM if computing resources are at all limited.
